---
title: "Control Flow Solutions"
author: "Julianne G. Clina"
institute: "University of Kansas Medical Center"
date: 2026-08-26
format:
  html:
    toc: true
    toc-depth: 2
    code-block-bg: true
    theme: cosmo
execute:
  echo: true
  eval: false
---

## Learning Objectives

By the end of this workshop, you should be able to:

- Write `if` / `else if` / `else` statements to classify a single value
- Use `switch()` to map a coded value to a label
- Write a `for` loop to repeat an action across multiple patients
- Use `next` to skip an iteration when data are missing

The core question to ask yourself throughout: **what decision am I making, and for how many things at once?** (`if` = one decision, one value at a time. `for` = repeat that decision across many values.)

```{r}
## Creating example dataset for practice!
library(tidyverse)

patients <- tibble(
  id = sprintf("P%02d", 1:12),
  age = c(34, 58, 45, 62, 29, 51, 47, 39, 66, 55, 42, 60),
  sbp = c(118, 142, 135, 128, 122, 158, NA, 131, 145, 126, 138, 165),
  activity_code = c(
    "A",
    "C",
    "B",
    "A",
    "D",
    "B",
    "C",
    "A",
    "B",
    "D",
    "C",
    "A"
  ),
  steps = c(
    8200,
    3100,
    5400,
    9100,
    11200,
    4800,
    6200,
    8900,
    5100,
    12400,
    4300,
    6700
  )
)

patients
```

`activity_code` is a stand-in for how activity level was recorded in the field: `"A"` = sedentary, `"B"` = lightly active, `"C"` = moderately active, `"D"` = very active.

---

## Activity 1: `if` / `else if` / `else`

### Why control flow?

Subsetting (last workshop) answers *"which rows meet a condition?"* Control flow answers a different question: *"given one value, which branch of logic should run?"* This matters when the action you want to take is more complex than just keeping or dropping a row — e.g., assigning a category, printing a message, or triggering different calculations.

The basic shape:

```{r}
# if (condition) {
# runs if condition is TRUE
#} else if (other_condition) {
# runs if condition is FALSE and other_condition is TRUE
#} else {
# runs if nothing above was TRUE
#}
```

A few things worth knowing:

- `if` expects a **single** `TRUE`/`FALSE` — not a vector. Feeding it a vector of length > 1 will error (or has in recent R versions).
- `if`/`else` returns a value, so you can assign the result directly: `x <- if (a) 1 else 2`.
- This is different from `ifelse()`, which *is* vectorized and works element-by-element over a whole vector — useful later, but not what we're practicing here.

### Step 1: Classify one patient's blood pressure

Standard categories (simplified for this exercise):

- `sbp < 120`: "Normal"
- `sbp >= 120 & sbp < 130`: "Elevated"
- `sbp >= 130 & sbp < 140`: "High Stage 1"
- `sbp >= 140`: "High Stage 2"

```{r}
# Using patients$sbp[2], write an if / else if / else chain
# that prints the correct category

sbp_val <- patients$sbp[2]

if (sbp_val < 120) {
  print("Normal")
} else if (sbp_val < 130) {
  print("Elevated")
} else if (sbp_val < 140) {
  print("High Stage 1")
} else {
  print("High Stage 2")
}
```

### Step 2: Handle missing data

Patient 7 has a missing `sbp` value. Right now, running your code from Step 1 on `patients$sbp[7]` will likely throw an error or behave oddly, because `if` can't evaluate `NA` as TRUE/FALSE.

```{r}
# Add a check at the top of your if/else chain: if the value is NA,
# print "Missing" instead of evaluating the BP conditions

sbp_val <- patients$sbp[7]

if (is.na(sbp_val)) {
  print("Missing")
} else if (sbp_val < 120) {
  print("Normal")
} else if (sbp_val < 130) {
  print("Elevated")
} else if (sbp_val < 140) {
  print("High Stage 1")
} else {
  print("High Stage 2")
}
```

**Why this matters for real data:** any time you write `if` around a variable that might have missing values, you need to decide up front what should happen for `NA` — R won't decide for you (it will error).

---

## Activity 2: `switch()`

`switch()` is a cleaner way to write an `if`/`else if` chain when you're matching one variable against a fixed set of known values — like a code.

```{r}
switch(
  "B",
  A = "Sedentary",
  B = "Lightly active",
  C = "Moderately active",
  D = "Very active"
)
```

Compare this to what the equivalent `if`/`else if` chain would look like — `switch()` says the same thing with a lot less repetition.

### Step 1: Translate one code

```{r}
# Use switch() to translate patients$activity_code[5] into its
# full label

switch(
  patients$activity_code[5],
  A = "Sedentary",
  B = "Lightly active",
  C = "Moderately active",
  D = "Very active"
)
```

### Step 2: Add a fallback

`switch()` can take an unnamed final argument as a default, for values that don't match anything listed:

```{r}
switch(
  "Z",
  A = "Sedentary",
  B = "Lightly active",
  "Unknown code"
)
```

```{r}
# Test your switch() from Step 1 on the value "Z" (a code that
# doesn't exist in our data) — add a default so it doesn't error

switch(
  "Z",
  A = "Sedentary",
  B = "Lightly active",
  C = "Moderately active",
  D = "Very active",
  "Unknown code"
)
```

---

## Activity 3: `for` loops

A `for` loop repeats the same block of code once per element of a sequence — useful when you want to apply an `if`/`else` decision (Activity 1) or a `switch()` (Activity 2) to *every* patient, not just one.

```{r}
for (i in 1:5) {
  print(i)
}
```

`i` is just a variable name — it takes on each value in `1:5` in turn, and the code inside `{}` runs once per value.

### Step 1: Loop through and classify every patient's BP

Combine what you built in Activity 1 with a `for` loop: loop over all 12 patients, and print each patient's `id` alongside their BP category. Don't forget to handle `NA` (patient 7).

```{r}
for (i in 1:nrow(patients)) {
  sbp_val <- patients$sbp[i]

  if (is.na(sbp_val)) {
    category <- "Missing"
  } else if (sbp_val < 120) {
    category <- "Normal"
  } else if (sbp_val < 130) {
    category <- "Elevated"
  } else if (sbp_val < 140) {
    category <- "High Stage 1"
  } else {
    category <- "High Stage 2"
  }

  print(paste(patients$id[i], "-", category))
}
```

### Step 2: Skip missing values with `next`

Instead of printing "Missing" for patient 7, use `next` to skip that iteration entirely — the loop moves on to the next patient without printing anything for the skipped one.

```{r}
for (i in 1:nrow(patients)) {
  sbp_val <- patients$sbp[i]

  if (is.na(sbp_val)) {
    next
  } else if (sbp_val < 120) {
    category <- "Normal"
  } else if (sbp_val < 130) {
    category <- "Elevated"
  } else if (sbp_val < 140) {
    category <- "High Stage 1"
  } else {
    category <- "High Stage 2"
  }

  print(paste(patients$id[i], "-", category))
}
```

### Step 3: Build a vector of results

Printing is fine for checking your work, but usually you want to **save** the result. Pre-allocate an empty character vector, then fill it in as the loop runs.

```{r}
bp_category <- vector("character", nrow(patients))

for (i in 1:nrow(patients)) {
  sbp_val <- patients$sbp[i]

  if (is.na(sbp_val)) {
    bp_category[i] <- NA
  } else if (sbp_val < 120) {
    bp_category[i] <- "Normal"
  } else if (sbp_val < 130) {
    bp_category[i] <- "Elevated"
  } else if (sbp_val < 140) {
    bp_category[i] <- "High Stage 1"
  } else {
    bp_category[i] <- "High Stage 2"
  }
}

patients$bp_category <- bp_category
patients
```

---

## Key Takeaways

1. **`if`/`else`** makes a decision for a *single* value — it can't take a vector.
2. **`switch()`** is a shortcut for matching one variable against a known list of codes, with an optional default.
3. **`for` loops** repeat a block of code once per element — this is how you scale a one-value decision (`if`/`switch`) up to a whole dataset.
4. **`next`** skips the rest of the current iteration and moves on — handy for skipping missing or bad data inside a loop.
5. Always ask: is this a single value (`if`/`switch`) or a whole vector/column at once (loop, or later: `ifelse()`/`purrr::map()`)?
